---
title: "STAA 551 Assignment 1, Spring 2020"
output:
  word_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE, echo=TRUE}
```


#### Name: Jon Dollard
####

1.  In R, load the data set "TreeCityUSA.csv", which we introduced in Lecture 1.  Use `dplyr` to select only the variables `fam_inc_med` and `pct_bachdeg` from this dataframe, and to change the missing values to NA. Give the name `data1` to this smaller dataframe.  

```{r}
#Install Tidyverse for ggplot2 capability
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
library(tidyverse)

#Access the dplyr library to cleanup the data and change -99 to NA
library(dplyr)

#Load the dataset TreeCityUSA to a dataframe called tree
tree <- read.csv("TreeCityUSA.csv")

#Cleanup fam_inc_med and pct_bachdeg missing data into R format
tree$fam_inc_med [tree$fam_inc_med ==-99] <- NA
tree$pct_bachdeg [tree$pct_bachdeg ==-99] <- NA

#Combine cleaned fam_inc_med and pct_bachdeg data into a single dataframe called data1
data1 <- select(tree, fam_inc_med, pct_bachdeg)
dim(data1)
summary(data1)
```
    
    (a)  Consider the manner in which the proportion of people with bachelor's degrees varies with median family income.  Obtain a scatter plot with `pct_bachdeg` on the vertical axis and `fam_inc_med` on the horizontal axis.  Give your plot a main title of "Problem 1(a)", and re-label the axes to something more informative.

```{r}
#Make a scatterplot of data1
ggplot(data = data1) + 
  geom_point(mapping = aes(x = fam_inc_med, y = pct_bachdeg)) +
  #geom_smooth(mapping = aes(x = fam_inc_med, y = pct_bachdeg),method=lm) +
  labs(
    title = "Problem 1(a)",
    x = "Median Income, USD ",
    y = "Proportion of Bachelors Degrees "
  ) 
```

    (b)  Based on the plot in (a), does there appear to be a relationship between the proportion of people with bachelor’s degrees and median family income? If so, what is it?

#Yes there appears to be a relationship between the proportion of people with a bachelor's degree and 
#median income.  The relationship appears to be that the given a higher median income the proportion of
#bachelor's degrees is also higher.

    (c)  Suppose we choose to consider a straight line model for this population.  Use the `lm` function in R to obtain the population regression equation.  State the population regression equation in proper format. 

```{r}
lm(data1$pct_bachdeg~data1$fam_inc_med)
#Regression equation:
#Let Y = pct_bachdeg
#Let X = fam_inc_med
```
$$ Y = -0.04458 + 0.000004509X + \epsilon $$
```

    (d)  Redo your scatter plot, this time with the population regression line superimposed. It would also be a good idea to graph a thicker line by including the `lwd=2` argument.  Comment on how the regression line does or does not fit the population.  

```{r}
ggplot(data = data1) + 
  geom_point(mapping = aes(x = fam_inc_med, y = pct_bachdeg)) +
  geom_smooth(mapping = aes(x = fam_inc_med, y = pct_bachdeg),method=lm, lwd = 2) +
  labs(
    title = "Problem 1(a)",
    x = "Median Income, USD ",
    y = "Proportion of Bachelors Degrees "
  ) 
#The regression line appears to fit the data for median incomes up to about $200,000.  For median
#incomes greater than $200,000 the regression line model does not appear to be a good fit.
```

2.  Obtain a sample of 30 observations from the data frame in Problem 1 (`data1`).  Give this dataframe the name `sample2`.

```{r}
#use dplyr sample_n function for random sample of 30 rows from data frame data1
sample2 = sample_n(data1, 30)
sample2
```

    (a)  Obtain a scatterplot similar to the one in 1(a), using the sample data.

```{r}
ggplot(data = sample2) + 
  geom_point(mapping = aes(x = fam_inc_med, y = pct_bachdeg)) +
  #geom_smooth(mapping = aes(x = fam_inc_med, y = pct_bachdeg),method=lm) +
  labs(
    title = "Problem 2(a)",
    x = "Median Income, USD ",
    y = "Proportion of Bachelors Degrees "
  ) 
```

    (b)  Would the scatterplot for this sample lead you to consider a straight line fit for this sample?  Why or why not? 

#Yes, from the scatterplot I would consider a straight line fit for this sample.  There appears
#to be a linear relationship between family median income and the proportion of bachelors degrees.

    (c)  Obtain the sample regression equation using the `lm` function in R. You will want to add the argument `data=sample2`, since these variables have the same name as in `data1`. State the sample regression equation in proper format.

```{r}
lm(sample2$pct_bachdeg~sample2$fam_inc_med)
#Regression equation:
#Let Y = pct_bachdeg
#Let X = fam_inc_med
#Y = -0.01808 + 0.000004140X
#Y = -0.04458 + 0.000004509X
```

    (d)  How similar are $b_0$ and $b_1$ to $\beta_0$ and $\beta_1$?  Would you expect them to be similar?  Why or why not? 

#I think that bo and b1 are somewhat similiar to beta-0 and beta-1, but their values are very dependent #on the actual sample results.  I would expect them to be simliar because it appears that a linear model #is reasonable for this data.

3.  Consider the data given in Exercise F on page 99 of Draper/Smith. The summary values for this sample are given.

```{r}
install.packages("aprean3", repos = "http://cran.us.r-project.org" )
library(aprean3)
attach(dse03f)
head(dse03f)
#dim(dse03f)
#dse03f
```

    (a) Using the summary values given, write the normal equations for this example.  Leave $b_0$ and $b_1$ as variables.  

#Normal Equation 1: 62 - 12*bo - 63.6*b1 = 0
#Normal Equation 2: 339.1 - 63.6*bo -339.18*b1 = 0

    (b) Solve the normal equations written in 3(a) algebraically for $b_0$ and $b_1$.  Include at least one intermediate step.
    
#b1 = (62 - 12*bo)/63.6
#339.1 - 63.6*bo - 339.18*((62 - 12*bo)/63.6) = 0
#After some rearranging and algebra
#-0.3962264*bo = 8.45283
#bo = -21.33
#b1 = 5

    (c)  Check your answers in (b) using the `lm` function in R.  You may want to load the data file from the `aprean3` package.  This dataset is given by `dse03f`.

```{r}
lm(dse03f$y~dse03f$x)
```

    (d)  Confirm numerically that the sample regression line passes through the point $(\bar{X},\bar{Y})$.
    
```{r}
X_bar = 5.3
Y_bar = 5.17
Y_bar_check = -21.33 + 5*X_bar
#This calculation confirms that the regression line passes through the point (X_bar, Y_bar)
```

    (e)  Confirm numerically that the residuals sum to zero.
    
```{r}
#vector of the values of X from dse03f
Xi = dse03f$x
#vector of sample predicted Y values
Y_hat = -21.33 + 5*Xi
#calculate the residuals
res = dse03f$y - Y_hat
#sum the residuals to confirm the sum is zero. 
s_res = sum(res)
s_res
#sum of the residuals is not exactly zero due to rounding error of the model
```

    (f)  Express the sample regression equation by pivoting around $(\bar{X},\bar{Y})$.

#Y_hat = Y_bar + b1*(X - X_bar)
#Y_hat = 5.17 + 5*(X - 5.3)

    (g)  Compute $SS_{reg}$, $SS_{res}$, $S_{YY}$, the corresponding degrees of freedom and mean squares, and display them in a basic ANOVA table.

```{r}
#Compute the SSreg
SSreg <- sum((Y_hat - mean(dse03f$y))^2)
SSreg
#compute the SSres
SSres <- sum((dse03f$y - Y_hat)^2)
SSres
#Compute the SSyy
SSyy <- sum((dse03f$y - mean(dse03f$y))^2)
SSyy
```

|   Source             |  df   |   Sum of Sq   |    Mean Sq    |    
|:---------------------|------:|--------------:|--------------:|
|  Due to b1|bo        |   1   |     52.5      |    52.5       | 
|  Residual            |  10   |     17.17     |   1.717       | 
|----------------------|-------|---------------|---------------|     
|  Total Corrected     |  11   |     69.67     |               |
  

    (h)  Compute $R^2$ and give an interpretation of its value.

```{r}
Rsqd = SSreg/SSyy
Rsqd
#75% of the total variation of the sample data about the mean of Y can be explained by the regression.
```

    (i)  Compute $SS(b_0)$, $SS(b_1|b_0)$, $\sum{Y^2}$, the appropriate degrees of freedom and mean squares, and display them in an "extra sum of squares" ANOVA table.

```{r}
#Compute the SSbo
n = 12
SSbo = n*(mean(dse03f$y)^2)
SSbo

#Compute the SSb1|bo
SSb1bo = sum((Y_hat - mean(dse03f$y))^2)
SSb1bo

#Compute the sum Y^2
sum_Ysq = sum(dse03f$y^2)
sum_Ysq
```

|   Source             |  df   |   Sum of Sq   |    Mean Sq    |    
|:---------------------|------:|--------------:|--------------:|
|  Due to b1|bo        |   1   |     52.5      |    52.5       | 
|  Residual            |  10   |     17.17     |   1.717       | 
|----------------------|-------|---------------|---------------|     
|  Total Corrected     |  11   |     69.67     |               |
| Correction Factor    |       |               |               |
|   (due to bo)        |   1   |     320.333   |               |
|----------------------|-------|---------------|---------------|
|  Total               |  12   |               |               |  


4.  For Problem 3, we consider $\beta_1$ to indicate the effect that mix moisture has on product density.  In other words, if $\beta_1$ is not zero, then it is a measure of the amount of change in product density for a one unit increase in mix moisture.  Use your calculations from Problem 3 to complete the following:

    (a) Plot the sample regression equation on a scatterplot of the two variables and comment on how well the regression equation describes the data.

```{r}
#4(a)
ggplot(data = dse03f) + 
  geom_point(mapping = aes(x = x, y = y)) +
  geom_smooth(mapping = aes(x = x, y = y), method=lm, se = FALSE) +
  labs(
    title = "Problem 4(a)",
    x = "Mix Moisture",
    y = "Density"
  ) 

#The regression appears to describe the data fairly well.  The data appears to be evenly distributed
#on either side of the regression line with no apparent outliers.
```

    (b) Compute a 99% confidence interval for $\beta_1$. 

```{r}
alpha = .01
b1 = 5
n = 12
MSres = SSres/(n-2)
Sxx = sum((dse03f$x - mean(dse03f$x))^2)
tval = qt(1-alpha/2,n-2)
s_b1 = sqrt(MSres/Sxx)
lower = b1 - tval*s_b1
lower
upper = b1 + tval*s_b1
upper
```

    (c) Give an interpretation of the confidence interval obtained in (b), in terms of the wording of this problem (i.e. moisture content and product density).  Do not associate your confidence level with a *probability*! 
    
##If we sampled this data many times we would find that 99% of the time b1 would be contained within
#the interval [2.135,7.865].

    (d) Consider the following statement: *"Changing the moisture content of a mix has **no** effect on its finished product density."*  Based on your confidence interval, would you tend to agree or disagree with that statement?  Why, or why not? 

#Based on the calculated CI in part (b) I would tend to disagree with that statement.  Since the CI does
#not contain zero there is evidence that changing the moisture content of a mix does have an effect
#on its finished product density.

    (e) Perform a t-test to test $H_0: \beta_1 = 0$ at the .01 significance level.  Include all of the steps indicated in lecture.  Note that your conclusion for this test must agree with your conclusion from the confidence interval!
  
#step 1
#Define the hypothesis to be tested
#Ho: Beta_1 = 0 vs. Ha: Beta_1 != 0

#Step 2
#Set significance level
alpha = 0.01

#Step 3
#State assumptions
#Assume that b1 is distributed N(Beta_1, sigma^2/Sxx)

#Step 4
#State the decision rule
#Reject Ho when |t| > t(1-alpha,n-2)
tval = qt(1-alpha/2,n-2)
tval

#Step 5
#Calculate T statistic from the sample
t_calc = b1/s_b1
t_calc

#Step 6
#Calculate the p value
p_value = 2*pt(abs(t_calc),n-2,lower.tail=FALSE)
p_value

#Step 7
#State Conclusion
#Given that our rejection criteria was met we will reject Ho and conclude that we have evidence that
#beta_1 is not equal to zero, and a linear relationship does exist between the moisture content of a 
#mix and the finished product density.  
    
    (f) Perform an F-test to test $H_0: \beta_1 = 0$ at the .01 significance level.  Include all of the steps indicated in lecture.  Note that your p-value and conclusion from this test should match the p-value and conclusion from (e).

#step 1
#Define the hypothesis to be tested
#Ho: Beta_1 = 0 vs. Ha: Beta_1 != 0

#Step 2
#Set significance level
alpha = 0.01

#Step 3
#State assumptions
#Assume that b1 is distributed N(Beta_1, sigma^2/Sxx)

#Step 4
#State the decision rule
df1 = 1
df2 = n-2
#Reject Ho when |f_calc| > f(1-alpha, df1, df2)
Fval = qf(1-alpha/2,df1, df2)
Fval

#Step 5
#Calculate F statistic from the sample
MSreg = SSreg
f_calc = MSreg/MSres
f_calc

#Step 6
#Calculate the p value
p_value = pf(f_calc,df1,df2,lower.tail=FALSE)
p_value

#Step 7
#State Conclusion
#Given that our rejection criteria was met we will reject Ho and conclude that we have evidence that
#beta_1 is not equal to zero, and a linear relationship does exist between the moisture content of a 
#mix and the finished product density.  (Same conclusion as part (e))
    
    (g) Show numerically for this data that the square of the computed t-statistic in (e) is equal to the computed F-statistic from (f). 

```{r}
t_squared = t_calc^2
t_squared
f_calc
#t_calc squared and f_calc are indeed equal 
```
    
    (h) Which of the procedures in (b), (e) or (f) do you find to be the most helpful, and why?  [This is your opinion!]

#I think my preference would be the t-test methodology.  The t distribution is easy for me to visualize
#and I feel comfortable with it.  Because it is symmetric it is easy to visualize the rejection regions
#and criteria.  In general I like the hypothesis testing methods since they provide a clear and logical #way to make a decision.  The t test is just a bit more familiar and comfortable to me than the F test. 
    
    (i) Compute the value of the correlation between product density and moisture content using the sums of squares and cross products from the textbook. Confirm numerically that the correlation is the square root of $R^2$ obtained in Problem 3.

```{r}
Syy = sum((dse03f$y - mean(dse03f$y))^2)
Sxy = sum((dse03f$x- mean(dse03f$x))*(dse03f$y-mean(dse03f$y)))
r = Sxy/sqrt(Sxx*Syy)
r
#confirm from problem 3
r_prob3 = sqrt(Rsqd)
r_prob3
```
    
